---
title: "README"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.width = 4, fig.height = 3, out.width = "50%")
source("Logistic_growth_model.R")
source("Predictability_of_model.R")
source("Integer_realization_of_mean_growth.R")
```

## Discrete time logistic model for population growth.

#### Notations:

- $N_t$ : Population size at time $t$
- $B_t$ : Total number of births in interval $(t, t+1)$
- $D_t$ : Total number of deaths in interval $(t, t+1)$
  - In this model $D_t = N_t$.
- $b_t = B_t / N_t$ := Per capita birth rate (:= expected litter size
  per individual ; see last section)
- $d_t = D_t / N_t$ := Per capita death rate
- $R_t = 1 + b_t - d_t$ := Growth rate
- $X_t = N_t/K$
- Note: $N_t,  B_t, D_t, b_t, R_t, X_t > 0$
- ***NBS*** := No biological sense.

------------------------------------------------------------------------

This is one of the simplest(!) model due to many underlying assumptions.

#### Assumptions:

- Time is discrete. and T is a time step.

  - We only know population size at time $0, T, 2T...$
  - We denote $t+T$ by $t+1$, and $t = 4$ means time $4T$.

- Non overlapping generations.

  - Each time step $T$ is a generation time.
  - i.e. All individual at time $t$, reproduce and immediately die at
    $t+T$.
  - No withing step reproduction of off springs.
  - So individual at $t$ won't be counted again at $t+T$ ; $d_t = 1$

- Population is Homogeneous.

  - No age, sex, spatial structures.
  - All offspring survives and reproduce with same expected fecundity.

- Population is closed.

  - Population change occur only due to death and births.
  - No immigration, no emigration.

- Logistic growth

  - Growth rate $R_t$ decrease linearly with time.

- **Mean field**

  - $N_t$ is very large \~(So we can apply law of large numbers to make
    underlying distribution structure immaterial).\~

  <!-- -->

  - $N_t$ represents **expected** population at time $t$, so
    $N_t \in \mathbb{R}_{\ge 0}$

  - Similarly $R_t$ is **expected** per capita growth rate at time $t$.
    and $R_t \in \mathbb{R}_{\ge 0}$

  - See last section
    `Stochasticity and real world interpretations of the model` for more
    details about underlying distributions and interpretations.

------------------------------------------------------------------------

#### Logistic model:

If we assume there are limited resources in the closed ecosystem, then
larger the population, more the fights for resources; resulting in the
high number of deaths and less number of births. This indicates that the
growth rate is a decreasing function of population size.

Let's assume simple linear relation between $R_t$ and $N_t$:
$$  R_t = R_m (1 - N_t/K) $$ Here, at $N_t = 0$, $R_t$ attains the
non-negative maxima denoted by $R_m$ and in this model N_t can not grow
beyond a point when $R_t = 0$, here $K$ is maximum value $N_t$ can
attain.

Therefore we get the following population growth equation:

$$N_{t+1} = R_m N_t (1 - N_t / K) $$ We divide both side by $K$ to
normalize $N_t$ to bound it in the interval $[0, 1]$. And let
$X_t = N_t/ K$ : = Normalized population. So, we get:
$$X_{t+1} = R_m X_t (1-X_t)$$

Note :

- For this model we have to restrict $0 \le R_m \le 4$ otherwise
  $0 \le X_t \le 1$ may not hold for all $t$, and hence ***NBS***.

- From assumption of Non overlapping generations we have $b_t = R_t$

- Due to mean field assumption we don't have to restrict $N_t$ to
  $\mathbb{Z}_{\ge 0}$ for sake of biological meaning.

------------------------------------------------------------------------

#### Example Trajectories:

- you can play with values of R_m to check for different R_m.

```{r fig.show="hold"}
plot_each_R_m(R_m_list = c(0.5, 1.5, 2.5, 3.2, 3.6, 4))
```

#### Observation:

- For $R_m \le 1$, $X_t \rightarrow 0$, $\forall X_0$.
- For $R_m < 3$ $X_t$ converges to $X^*$, $\forall X_0$ except
  $X_0 = 0, 1$,
- For $3 \lt R_m \lt 3.44$ population oscillates between 2 points.
- For $3.45 \lt R_m \lt 3.5$ oscillates between 4 points.
- For $3.5 \lt R_m$ it become chaotic and unpredictable.

------------------------------------------------------------------------

#### Equilibrium/ fixed point : $X^*$

Population size, if attained, stays constant over all the following
generations.

> Let **E** be state space consisting of all population sizes, and **f**
> be evolution function (i.e. flow) then $X_{t+1} = f(X_t)$. And $X^*$
> is a fixed point/equilibrium if $X^* = f(X^*)$.

$$\begin{aligned}
& X^*= R_m X^*(1 - X^*) \\
& \\
\text{Positive Solutions:}\\
& \text{a) } X^* = 0  & \text{for } R_m \in [ 0, 4] \\
& \text{b) } X^* = 1 - \frac{1}{R_m} & \text{for } R_m \in [1, 4]
\end{aligned}$$

```{r, fig.show='hold'}
plot_eqb_curve()
```

------------------------------------------------------------------------

### Stability of equilibrium:

Small perturbation to $X^*$ stays small and eventually dies out over
time.

> [**Asymptotically stable:**]{.underline}
>
> Let $X^*$ be fixed point of **f**, if $\forall \epsilon \gt 0$ ,
> $\exists \delta \gt 0$ s.t.
> $$ |X_0 - X^* | \lt \delta \implies |X_n - X^*| \lt \epsilon, \forall n \ge 0 $$
>
> and $\exists \delta_0 \gt 0$ s.t
> $$|X_0 - X^* | \lt \delta_0 \implies X_n \rightarrow X^*$$

> [**Linear stability test:**]{.underline}
>
> Let $f : E \rightarrow E$ be evolution function where $E$ is
> $\mathbb{R}^+$. And let $f$ be $C^1$ near fixed point $X^*$, then
> $$\begin{aligned} |f'( X^* ) | \lt 1 & \implies X^* \text{ is asymptotically stable. } \\ |f'(X^) | \gt 1 & \implies X^* \text{ is unstable.} \end{aligned} $$

#### Observations:

- For $R_m \le 1$ we have stable equilibrium at $0$.
- For $R_m \in (1, 2]$ we have unstable equilibrium at $0$.
- For $1 \lt R_m \lt 3$ gives stable equilibrium at $(1 - 1/R_m)$.
- For $R_m = 3$, stability undetermined.

------------------------------------------------------------------------

#### Bifurcation:

> [**Attractor points of**$\{X_n\}$ :]{.underline} we plot subsequential
> limit points of sequence of population size time series $X_n$. for
> given $R_m$ set of attractor points could be single point where the
> orbit converges, finitely many points on a periodic cycle, or a
> continuum/Cantor-like set under chaos.

> [**Bifurcation diagram (discrete case)**]{.underline} : For each $R_m$
> we plotted last $200$ values of $X_t$. i.e.values from
> $\phi_{800}(0.1; R_m)$ to $\phi_{1000}(0.1; R_m)$ against each $R_m$.
>
> - you can play with values of $x_0$

```{r}
Plot_Bifurcation(x_0 = 0.1)
```

------------------------------------------------------------------------

#### Determinism, Robustness

- Logistic model is completely deterministic. No stochastic elements
  involved (`This is not so maybe concluded from last section)`.

- Model is robust except at the bifurcation points. (trivial)

  - Robustness (or its loss, at a bifurcation) is what early-warning
    signals detect.

------------------------------------------------------------------------

#### Predictability

Check `predictability of model.R` for simulation.

- we have computationally checked the for error values
  $\delta(0) = 0.1, 0.01, 0.00001$

  - you can check for multiple values by just changing `error` vector in
    the function input.

  - `Plot_error_trajectories(dynamic function, initial states, parameter values, time step, error vector)` -
    you can change input variables and play around with any discrete
    time dynamic function to see how they respond to errors.

- Evolution of error in discrete logistic model:

  - here $\delta(t) \text{ is relabeld as } E_t$

```{r fig.show="hold"}
Plot_error_trajectories(X_0_list = c(0.1), R_m_list = c(0.5, 1.5, 2.5, 3.2, 3.6, 4), t = 1:100, eps = c(0.1, 0.01, 0.00001))
```

#### Observation:

- For $R_m \lt 3$ we have stable equilibrium i.e. fixed point Attractor
  and hence population growth is completely predictable.
  ($\lambda \lt 0$) and basin of attractor is whole set $(0, 1)$ except
  $x_0 = 0, 1$.

- For $3 \lt R_m \lt 3.5$ we have oscillating population i.e limit cycle
  attractor and we can see that errors are showing phase drift, so
  population growth is predictable up to phase drift $(\lambda = 0)$

- For $R_m >3$ we have chaos in the system and population is
  unpredictable.

- Note: For $R_m = 3.6$ in graph of $E_t$, Error trajectory of
  $\delta(0) = 0.00001$`red` is near zero for $t \lt \approx 25$ ; which
  can be seen as a predictability time horizon where error
  $E_t \lt \varepsilon_\text{tol}$

------------------------------------------------------------------------

### Stochasticity and real world interpretation of the model:

- Whenever we take $R_t$ linearly decreasing with increasing $N_t$,
  values of $R_t$ lands on many fractional numbers or even irrational
  numbers (but computationally impossible). These fractional values of
  $R_t$ invokes non-integer population sizes; which makes No biological
  sense as population must be an integer. And therefore we need to
  summon randomness to handle it.

- We can not avoid randomness even when formulation is completely
  deterministic.

- Let per capita birth rate $b_t = 3$ ; think about what does it mean?
  Each individual gives $3$ litters between $(t, t+1)$ ? But even when
  half population have $6$ litters and half have none, we still get
  $b_t = 3$. So, we can have different distribution for given $b_t$ or
  even $N_t$ for that matter.

- Let's consider strictly integer population size
  $N_t \in \mathbb{Z}_{\ge 0}$ with per capita birth rate
  $b_t \in \mathbb{R}$ and suppose each individual $i$ may have any
  number of litter size. So Let $\Omega_i = \{0, 1, 2,... \}$ be that
  sample space and $X_i : \Omega_i \to \mathbb{R}$ be
  $\text{Poisson}(b_t)$ random variable denoting Numbers of litter of
  $i$-th individual (you may choose any other distribution, Negative
  binomial is a good choice as it's most close to reality). Here $b_t$
  can be interpret as expected litter size for each individual. I.e. if
  you re-run the experiment with same conditions for large number of
  times then the average the litter size = $b_t$ and probability of
  having $k$ litter is given by PMF
  $P(X_i = k) = e^{-b_t} \frac{b_t^k}{k!}$, further we let
  $S = \sum_{i =1}^{N_t} X_i$ be random variable denoting total Number
  of births in population and since $\{X_i\}$ are iid
  $\text{Poisson}(b_t)$, $S \sim \text{Poisson}(N_t \cdot b_t)$. And
  since $d_t = 1$, $N_{t+1} = S$ , So $N_{t+1}$ is a random variable,
  that takes strictly integer values; with poisson distribution and
  expected value $E(N_{t+1}) =N_t\cdot b_t$.

- To make our life simple, we take assumption that $N_t$ **is large,\
  So We can summon LLN and therefore**
  $$N_{t+1} \approx E(N_{t+1}) =N_t\cdot b_t$$

  - I.e. realized per-capita growth rate converges to the parameter
    $b_t$. This justifies that stochastic branching process can be
    treated as deterministic recursion $N_{t+1}=N_tb_t$.
  - No matter what underlying distributions is as long as $X_i$ are
    independent, from central limit theorem we recover the normal
    distribution as $N_t$ is large. So deterministic model still works
    for large $N_t$.

- Note : Near extinction points, when $N_t$ is small, $N_{t+1}=N_tb_t$
  is no longer even approximately valid and we need to treat
  stochasticity carefully.

  In `Integer_realization_of_mean_growth.R` I tried to create logistic
  model with taking $N_t$ as integer values by realizing it using
  different well known distributions to see how population dynamics
  works. Here are some graphs for
  $N_{t+1} \sim \text{Poisson}(R_t \cdot N_t)$. You can play with values
  and distributions {"poisson", "negbinom", "normal", "none"}.

  Note: for 'negbinom', i guess due to high dispersion value which is
  set at 1 in the code, and since population can no recover from 0,
  graphs eventually end up at 0

```{r}
plot_each_R_m_int(N_0_list = c(100, 200, 500, 800), R_m_list = c(0.5, 1.5, 2.5, 3.2,3.6, 3.9), K = 1000, time_steps = 100, distribution = "poisson")
```

```{r Stocastic Bifurcation}
Plot_Bifurcation_int(N_0 = 100, R_m_list = seq(0, 3.9, 0.01), K = 1000, time_steps = 1000, distribution = "poisson")
```

------------------------------------------------------------------------

------------------------------------------------------------------------
